---
title: "clim_data_copernicus"
author: "Shahar Chaikin"
date: "2025-09-09"
output: html_document
---

Here we will extract climatic data including SST and SBT based on the Copernicus Marine Service and the 'CopernicusDownloadR' R package

Packages
```{r}
library(CopernicusDownloadeR)
library(h3)
```

Extract the h3 IDs at resoultion 3 for the MEditerranean Sea
```{r}
# Bounding box for the Mediterranean Sea
min_lat <- 30.26
max_lat <- 45.78
min_lon <- -6.03
max_lon <- 36.22

```

# Get H3 Polygons within the Bounding Box
1) Define a set of points that cover the bounding box. A simple way is to create a grid.
2) Convert these points to H3 indexes at your desired resolution (3 in this case).
3) Get the parent polygons of these indexes. This ensures you have the H3 hexagons that are exactly at resolution 3 and contain your grid points.
```{r}
# Create a grid of points within the bounding box
lat_grid <- seq(min_lat, max_lat, by = 0.5)
lon_grid <- seq(min_lon, max_lon, by = 0.5)
grid_points <- expand.grid(lat = lat_grid, lon = lon_grid)

# Convert each grid point to an H3 index at resolution 3
h3_indexes <- h3js::h3_geo_to_h3(grid_points$lat, grid_points$lon, res = 3)

# Get the unique parent H3 indexes at resolution 3
h3_polygons <- unique(h3_indexes)
```

More accurate bbox
```{r}
# Get coastline data from rnaturalearth
coastlines <-rnaturalearth::ne_coastline(scale = "large", returnclass = "sf")
med_shape=sf::st_read("C:\\Users\\User\\Desktop\\research\\Mentoring and colaborations\\Georgous\\MHW_NIS\\World_Seas_IHO_v3\\iho.shp")

# Combine all polygons in the shapefile into a single polygon
med_sea_polygon <- sf::st_union(med_shape)
str(med_sea_polygon)
plot(med_sea_polygon[2])
# Use h3_polyfill to get all H3 IDs at resolution 3 that intersect with the polygon
med_h3_ids <-h3jsr::polygon_to_cells(med_sea_polygon[2], res = 3)
med_h3_ids_vec=unlist(med_h3_ids) %>% tibble::as.tibble()
#Some are missing
missing=c("831edcfffffffff",
          "831ed1fffffffff",
          "831edefffffffff",
          "831edafffffffff",
          "833f72fffffffff",
          "833f73fffffffff",
          "833f29fffffffff",
          "832db0fffffffff",
          "832db1fffffffff",
          "832da3fffffffff",
          "833f68fffffffff",
          "831ecafffffffff",
          "831ecefffffffff",
          "831ec3fffffffff",
          "833e2cfffffffff",
          "833f5afffffffff",
          "833e2dfffffffff",
          "833e25fffffffff",
          "833f19fffffffff",
          "833f69fffffffff",
          "833f63fffffffff",
          "833f64fffffffff",
          "833f2cfffffffff",
          "831e8bfffffffff",
          "833f24fffffffff",
          "833f26fffffffff",
          "831e8afffffffff",
          "831ea3fffffffff",
          "831ea5fffffffff",
          "831ea0fffffffff",
          "833facfffffffff",
          "833f12fffffffff",
          "83384afffffffff",
          "833968fffffffff",
          "833f34fffffffff",
          "831e85fffffffff",
          "831e81fffffffff",
          "833954fffffffff",
          "833973fffffffff",
          "833835fffffffff",
          "833834fffffffff",
          "83391afffffffff",
          "831e94fffffffff",
          "831e96fffffffff",
          "831eb0fffffffff",
          "831f9bfffffffff",
          "833919fffffffff",
          "833918fffffffff",
          "833824fffffffff",
          "833823fffffffff",
          "83382efffffffff",
          "833863fffffffff",
          "83386afffffffff",
          "83386efffffffff",
          "833875fffffffff",
          "833862fffffffff") %>% 
  tibble::as.tibble()

#add missing
med_h3_ids_vec=med_h3_ids_vec %>% 
  rbind(missing)

write.csv(med_h3_ids_vec,"h3_polygons.csv",row.names = F)
```



# setup the environment using Copernicus log in credentials
```{r}
cm <- setupCopernicusEnvironment("schaikin", "Aa123456")
```

# list available variables names - get complete variable information
```{r}
csv_path="C:\\Users\\User\\AppData\\Local\\R\\win-library\\4.3\\CopernicusDownloadeR\\extdata\\CopernicusVariables&Datasets.csv"

listVariables (show="completeList",csv_path =csv_path)

listVariables (csv_path = csv_path)
```

# Example of using the function to extract one variable for specific location point 
```{r}
copernicusData <- extractData(
  cm = cm,  # Copernicus Marine object from setup function
  variables = "oxygen",
  lon = -9.64671906997131, lat = 36.7143279919028,
  startDate = "2021-01-01", endDate = "2021-01-20",
  depth = c(0,11),
  frequency = "daily",
  data_output = "path_to_data_folder",
  csv_path = csv_path)  # Path to save data
```

# Example of using the function to extract multiple variables, large geographic area, while  selecting download path
```{r}
copernicusData <- extractData(
  cm = cm,  # Copernicus Marine object from setup function
  variables = c("oceanTemperature", "oxygen"),
  lon =c(-30,10),
  lat = c(30,45),
  startDate = "2020-03-01", endDate = "2020-06-20",
  depth = c(1), 
  frequency = "monthly",
  data_output = "path_to_data_folder" )  # Path to save data 
```

# visualise downloaded data
```{r}
library(ncdf4)
library(terra)
```

# Set the path for the NetCDF file
# Replace "path_to_data_folder" with the file path to your downloaded dataset
```{r}
ncfile <- "C:\\Users\\User\\Desktop\\research\\Mentoring and colaborations\\Georgous\\MHW_NIS\\path_to_data_folder\\cmems_mod_glo_bgc_my_0.25deg_P1D-m_o2_9.75W_36.75N_0.51-9.82m_2021-01-01-2021-01-20.nc"
```

# Open the NetCDF file
```{r}
nc <- nc_open(ncfile)
# Print information about the NetCDF file
print(nc)
#Import a multi-band NetCDF file
multi <- rast(ncfile)
# Print a summary of the multi-band raster
print(multi)
# Plot the multi band raster
plot(multi)
# Plot the first band (e.g., depth level or first time step)
plot(multi[[2]], main = "Multi-Band NetCDF Raster: Band 2")
# save selected bands data as raster 
writeRaster(multi[[2]], filename = "o2_test.tif", overwrite = TRUE)
```


# DATA POINTS
```{r}
coords <- read.csv2("./coords.csv",sep=",",h=F)
colnames(coords) <- c("country","HEX")
#coords <- read.csv2("./h3_ids.csv")
new_h3 <- read.csv2("C:\\Users\\User\\Desktop\\research\\postdoc\\Araujo\\MHWs_project\\R\\MHW_postdoc_project\\h3_ids.csv")
coords <- data.frame(HEX = new_h3$h3_id)

pol <-h3jsr::cell_to_polygon(coords$HEX)
h3_range <- as.data.frame(do.call("rbind", lapply(sf::st_geometry(pol), sf::st_bbox)))

nrow(h3_range) == nrow(coords)
```

# Get Data
```{r}
for (coord in 1:nrow(coords)){
  
  print(coords[coord,])
  lat_r <- sort(c(h3_range$ymin[coord],h3_range$ymax[coord]))
  lon_r <- sort(c(h3_range$xmin[coord],h3_range$xmax[coord]))
  
  copernicusData <- extractData(
    cm = cm,  # Copernicus Marine object from setup function
    variables = c("oceanTemperature"),
    lon =lon_r,
    lat = lat_r,
    startDate = "1970-01-01", endDate = "2022-01-01",
    depth = c(0,1800), 
    frequency = "daily",
    data_output = "./Daily_Data/" )  # Path to save data 
  
    files <- list.files("./Daily_Data/", pattern = "\\.csv$", full.names = TRUE)
    files_info <- file.info(files)
    csv_n <- files[which.max(files_info$mtime)]
    
    print(paste("Archivo más reciente:", csv_n))
    
    csv_i <- read.csv2(csv_n,sep=",")
    csv_i <- cbind(coords[coord,],csv_i)
    write.csv2(csv_i,csv_n)
}
```