---
title: "stage_4_bind_mhw_records"
author: "Shahar Chaikin"
date: "2025-09-24"
output: html_document
---

Bind climatologies with records data

Libraries
```{r}
library(tidyverse)
```

data
```{r}
#trended
clim=read_rds("climatology_df_median_sst_10m.rds")
events=read_rds("event_df_median_sst_10m.rds")
#detrended
clim_dt=read_rds("climatology_dt_df.rds")
events_dt=read_rds("event_dt_df.rds")
```

#1.1) annual cumulative intensity
Estimate annual cumulative intensity per hexagon
```{r}
annual_c_int=clim %>% 
  drop_na(median_sst_10m) %>% 
  mutate(
    # Create a new column, e.g., 'temp_anomaly'
    intensity = dplyr::if_else(
      !is.na(event_no), # The condition: is event_no not NA?
      median_sst_10m - seas, # What to do if the condition is TRUE
      0 # What to do if the condition is FALSE
    )) %>% 
    mutate(year_of_mhws=lubridate::year(date)) %>% 
  group_by(h3_id,year_of_mhws) %>% 
  summarise(annual_c_int=sum(intensity)) %>% 
  mutate(year_of_records=year_of_mhws+1)
```

#1.2) Records data
Add records data
```{r}
records=read_csv("records_cleaned.csv") %>% 
  mutate(
    h3_id = h3jsr::point_to_cell(
      input = sf::st_as_sf(., coords = c("long", "lat"), crs = 4326),
      res = 3)) %>% 
  #remove marmara and black sea records that are not in the climatic model
  filter(!h3_id%in%c("831ec8fffffffff", "831ec9fffffffff", "832d05fffffffff"))

#For every h3_id, year,lat, and long keep only the first record
records_min=records %>% 
  group_by(h3_id,species) %>% 
  filter(year%in%min(year)) %>% 
 #For any remaining groups (h3_id/species) with multiple records in the minimum year, keep only the first one
  slice(1) %>%
#Ungroup the data (good practice after grouping operations)
  ungroup()
#Sum annual and spatial records
records_min_count=records_min %>% 
  group_by(h3_id,year) %>% 
  summarise(records=n()) %>% 
  rename(year_of_records=year)
#plot
ggplot(records_min_count)+
  geom_point(aes(x=year_of_records,y=records))
```

##1.2.1) Plot records
Plot coords
```{r}
# Assuming your dataframe is named 'records'
# Ensure 'lat' and 'long' columns exist in 'records'

# 1. Get world map data
world_map <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

# 2. Convert your 'records' data frame to an sf object
records_sf <- records %>%
  drop_na(lat, long) %>% # Remove rows with missing lat/long
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) # CRS 4326 is WGS84 for lat/long

# 3. Calculate the extent of your coordinates
bbox <- sf::st_bbox(records_sf)

# 4. Create the ggplot
ggplot() +
  # Add the world map
  geom_sf(data = world_map, fill = "lightgrey", color = "darkgrey") +
  # Add your data points
  geom_sf(data = records_sf, size = 1, alpha = 0.7) +
  # Crop the map to the extent of your points
  coord_sf(
    xlim = c(bbox["xmin"] - 1, bbox["xmax"] + 1), # Adjust buffer as needed
    ylim = c(bbox["ymin"] - 1, bbox["ymax"] + 1), # Adjust buffer as needed
    expand = FALSE # Prevents adding extra space around the crop
  ) +
  # Add labels and theme
  labs(
    title = "Species records",
    x = "Longitude",
    y = "Latitude",
    color = "Species" # Legend title for species color
  ) +
  theme_minimal()+
  guides(color=F)
```

#1.3) Bind MHWs and records
Bind MHWs with the records of the succeeding year
```{r}
c_int_records=annual_c_int %>% 
  left_join(records_min_count,
            by=c("h3_id","year_of_records")) %>% 
  mutate(records = replace_na(records, 0))
#write.csv(c_int_records,"c_int_records_pseudo_first.csv",row.names = F)
```

Explore
```{r}
ggplot(c_int_records)+
  geom_point(aes(x=year_of_records,y=records))

ggplot(c_int_records)+
  geom_point(aes(x=annual_c_int,y=records))

ggplot(c_int_records)+
  geom_point(aes(x=year_of_records,y=annual_c_int))
```

#1.4) GLMMs
```{r}
library(glmmTMB)

#nbinom2
test2=glmmTMB(data = c_int_records,formula = records~annual_c_int+(1|h3_id),family = "nbinom2")
summary(test2)
sjPlot::plot_model(test2,type="eff",terms="annual_c_int",show.data = F)+
  theme_bw()
plot(DHARMa::simulateResiduals(test2))
MuMIn::r.squaredGLMM(test2)
AIC(test2)

#nbinom
test3=glmmTMB(data = c_int_records,formula = records~annual_c_int+(1|h3_id),family = "nbinom1")
summary(test3)
sjPlot::plot_model(test3,type="eff",terms="annual_c_int",show.data = F)
plot(DHARMa::simulateResiduals(test3))
MuMIn::r.squaredGLMM(test3)
AIC(test3)
```

#2.1) detrended cumulative intensity
Estimate annual cumulative intensity per hexagon
```{r}
annual_c_int_dt=clim_dt %>% 
  drop_na(residuals) %>% 
  mutate(
    # Create a new column, e.g., 'intensity'
    intensity = dplyr::if_else(
      !is.na(event_no), # The condition: is event_no not NA?
      residuals - seas, # What to do if the condition is TRUE
      0 # What to do if the condition is FALSE
    )) %>% 
    mutate(year_of_mhws=lubridate::year(date)) %>% 
  group_by(h3_id,year_of_mhws) %>% 
  summarise(annual_c_int=sum(intensity)) %>% 
  mutate(year_of_records=year_of_mhws+1)
```

#2.2) Bind MHWs and records
Bind MHWs with the records of the succeeding year
```{r}
c_int_records_dt=annual_c_int_dt %>% 
  left_join(records_min_count,
            by=c("h3_id","year_of_records")) %>% 
  mutate(records = replace_na(records, 0))
#write.csv(c_int_records_dt,"c_int_records_pseudo_first_dt.csv",row.names = F)
```

Explore
```{r}
ggplot(c_int_records_dt)+
  geom_point(aes(x=year_of_records,y=records))

ggplot(c_int_records_dt)+
  geom_point(aes(x=annual_c_int,y=records))

ggplot(c_int_records_dt)+
  geom_point(aes(x=year_of_records,y=annual_c_int))
```

#1.4) GLMMs
```{r}
library(glmmTMB)

#nbinom2
test2_dt=glmmTMB(data = c_int_records_dt,formula = records~annual_c_int+(1|h3_id),family = "nbinom2")
summary(test2_dt)
sjPlot::plot_model(test2_dt,type="eff",terms="annual_c_int",show.data = F)+
  theme_bw()
plot(DHARMa::simulateResiduals(test2_dt))
MuMIn::r.squaredGLMM(test2_dt)
AIC(test2_dt)
#nbinom
test3_dt=glmmTMB(data = c_int_records_dt,formula = records~annual_c_int+(1|h3_id),family = "nbinom1")
summary(test3_dt)
sjPlot::plot_model(test3_dt,type="eff",terms="annual_c_int",show.data = F)
plot(DHARMa::simulateResiduals(test3_dt))
MuMIn::r.squaredGLMM(test3_dt)
AIC(test3_dt)
```

Extract predicted of the best model
```{r}
test3_dt_gg=ggeffects::ggpredict(test3_dt,terms="annual_c_int[0:965 by=1]")
```

Plot
```{r}
# 1. Extract the p-value for 'annual_c_int'
# The p-value is in the fixed effects part of the model summary.
p_value <- summary(test3_dt)$coefficients$cond["annual_c_int", "Pr(>|z|)"]

# 2. Extract the sample size (number of observations)
N <- nobs(test3_dt)

# 3. Create a formatted label string
# Format the p-value to a specific number of decimal places (e.g., 3)
p_label <- paste0("p-value: ", format.pval(p_value, digits = 3, eps = 0.001))
N_label <- paste0("n = ", N)

# 4. Determine the position for the labels
# These coordinates are illustrative; adjust them to fit your plot data.
# Assuming x is from 0 to about 120 and y is on a log scale (say max 1000).
x_pos <- max(c_int_records_dt$annual_c_int, na.rm = TRUE) * 0.95 # Near top-right
y_pos_p <- max(c_int_records_dt$records, na.rm = TRUE) * 0.95 
y_pos_N <- max(c_int_records_dt$records, na.rm = TRUE) * 0.7 # Adjust vertical separation on log scale

# Convert y_pos to a single data frame for geom_text/annotate if preferred
stats_df <- data.frame(
  x = c(x_pos, x_pos),
  y = c(y_pos_p, y_pos_N),
  label = c(p_label, N_label)
)

#plot
ggplot()+
  geom_point(data=c_int_records_dt,
             aes(x=annual_c_int,y=records),
             color="grey",
             alpha=0.5)+
  geom_ribbon(data=test3_dt_gg,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3)+
  geom_line(data=test3_dt_gg,
            aes(x=x,y=predicted))+
  scale_fill_gradient(low = "grey70", high = "red", guide = "none") +
  theme_bw()+
  scale_y_log10()+
  labs(x="Annual cumulative intensity (°C-days)",
       y="New records within the analysis period") +
  
  # Add the labels using annotate() for fixed positions
  annotate("text", x = x_pos, y = y_pos_p, 
           label = p_label, 
           hjust = 1, vjust = 1, size = 4) +
  annotate("text", x = x_pos, y = y_pos_N, 
           label = N_label, 
           hjust = 1, vjust = 1, size = 4) 
```

