---
title: "stage_4_bind_mhw_records"
author: "Shahar Chaikin"
date: "2025-09-24"
output: html_document
---

Bind climatologies with records data

Libraries
```{r}
library(tidyverse)
library(glmmTMB)
library(glmm.hp)
```

#data
```{r}
#data for models
data_models=read_rds("record_data_for_glmms.rds")
#for compatibility with variance partition
data_models_pre_scaled=data_models %>% 
  mutate(
    first_rec_year_sc = scale(first_rec_year)[, 1],
    CPI_sc = scale(CPI)[, 1],
    sources_sc = scale(sources)[, 1],
    annual_c_int_sc = scale(annual_c_int)[, 1],
    year_of_mhws_sc = scale(year_of_mhws)[, 1],
    median_sst_sc = scale(median_sst)[, 1],
    max_s_sst_sc=scale(max_s_sst)[,1])
  
```

#Modeling
##GLMM
```{r}
test_main=glmmTMB(
  data = data_models,
  formula = records~
    scale(first_rec_year)+
    scale(CPI)+
    scale(sources)+
    scale(annual_c_int)+
    scale(year_of_mhws)+
    scale(median_sst)+
    scale(max_s_sst)+
    (1|h3_id)+
    (1|species),
  family = "binomial",
  control = glmmTMBControl(optimizer = optim,
                           optArgs = list(method = "BFGS",
                                          maxit = 10000)))
#Colinearity
##median_s_sst
 #            Term  VIF   VIF 95% CI Increased SE Tolerance Tolerance 95% CI
 #   scale(median_sst) 5.90 [5.77, 6.04]         2.43      0.17     [0.17, 0.17]
 # scale(median_s_sst) 6.03 [5.90, 6.18]         2.46      0.17     [0.16, 0.17]

##max_s_sst - no colinearity!

##min_s_sst - no colinearity!

##mean_s_sst
# Moderate Correlation
#               Term  VIF   VIF 95% CI Increased SE Tolerance Tolerance 95% CI
#  scale(median_sst) 5.36 [5.24, 5.48]         2.32      0.19     [0.18, 0.19]
#  scale(mean_s_sst) 5.44 [5.32, 5.57]         2.33      0.18     [0.18, 0.19]

#write_rds(test_sp_dt_time_sst_mon_eff,"test_sp_dt_time_sst_mon_eff_binomial_s_sst.rds")
# test_sp_dt_time_sst_mon_eff=read_rds("glmms\\test_sp_dt_time_sst_mon_eff_cpi/test_sp_dt_time_sst_mon_eff_cpi.rds")
summary(test_sp_dt_time_sst_mon_eff)
plot(DHARMa::simulateResiduals(test_sp_dt_time_sst_mon_eff))
performance::check_singularity(test_sp_dt_time_sst_mon_eff)
performance::check_convergence(test_sp_dt_time_sst_mon_eff)
performance::check_overdispersion(test_sp_dt_time_sst_mon_eff)
performance::check_collinearity(test_sp_dt_time_sst_mon_eff)
MuMIn::r.squaredGLMM(test_sp_dt_time_sst_mon_eff)
#Summarise table for the model
fixed_effects_df <- broom.mixed::tidy(test_sp_dt_time_sst_mon_eff, effects = "fixed") %>%
  # Select and rename columns to match the request
  select(
    Variable = term,
    Coefficient = estimate,
    `Std. Error` = std.error,
    `Z Value` = statistic,
    `P Value` = p.value
  ) %>%
  # Format numeric columns for clean presentation (e.g., 3 decimal places)
  mutate(
    Coefficient = round(Coefficient, 3),
    `Std. Error` = round(`Std. Error`, 3),
    `Z Value` = round(`Z Value`, 2),
    `P Value` = format.pval(`P Value`, digits = 3, eps = 0.0001)
  )
#random effects
random_effects_df <- broom.mixed::tidy(test_sp_dt_time_sst_mon_eff, effects = "ran_pars") 

#write.csv(fixed_effects_df,"fixed_effects_df_binomial.csv",row.names = F)

#Binomial model: AIC = 7687.6
#Betabinomial model: AIC = 7689.6
#Delta is 2 - models are quite similar - I am chossing base on r^2
```
R2 partition
```{r}
####################
#Variance partition#
#####################
#Run Hierarchical Variance Partitioning using glmm.hp
## The glmm.hp function calculates the Marginal R2 (R2m) and then decomposes
## it into the individual contribution of each fixed effect predictor using
## hierarchical partitioning (average shared variance method).
library(glmm.hp)
#semantics
data_to_use <- model_data_2020_clean %>%
  mutate(
    first_rec_year_sc = scale(first_rec_year)[, 1],
    CPI_sc = scale(CPI)[, 1],
    sources_sc = scale(sources)[, 1],
    annual_c_int_sc = scale(annual_c_int)[, 1],
    year_of_mhws_sc = scale(year_of_mhws)[, 1],
    median_sst_sc = scale(median_sst)[, 1],
    max_s_sst_sc=scale(max_s_sst)[,1])

#model
test_sp_dt_time_sst_mon_eff_sem=glmmTMB(
  data = data_to_use,
  formula = records~
    first_rec_year_sc+
    CPI_sc+
    sources_sc+
    annual_c_int_sc+
    year_of_mhws_sc+
    median_sst_sc+
    max_s_sst_sc+
    (1|h3_id)+
    (1|species),
  family = "binomial")#,
  #control = glmmTMBControl(optimizer = optim,
  #                         optArgs = list(method = "BFGS",
  #                                        maxit = 10000)))
                                          
summary(test_sp_dt_time_sst_mon_eff_sem)
performance::check_singularity(test_sp_dt_time_sst_mon_eff_sem)
performance::check_convergence(test_sp_dt_time_sst_mon_eff_sem)
performance::check_collinearity(test_sp_dt_time_sst_mon_eff_sem)
MuMIn::r.squaredGLMM(test_sp_dt_time_sst_mon_eff_sem)
DHARMa::simulateResiduals(test_sp_dt_time_sst_mon_eff_sem) %>% plot
performance::check_overdispersion(test_sp_dt_time_sst_mon_eff_sem)
#pred list
# iv=list(pred1="first_rec_year_sc",
#         pred2="CPI_sc",
#         pred3="sources_sc",
#         pred4="annual_c_int_sc",
#         pred5="year_of_mhws_sc",
#         pred6="median_sst_sc")

R2_partition_results <-glmm.hp(test_sp_dt_time_sst_mon_eff_sem,
                                #iv =iv,
                                type="R2") # Specify that I want adjusted R2 partitioning

R2_partition_results
write_rds(R2_partition_results,"R2_partition_results_binomial_spatial_sst.rds")

#Interpretation
##Unique - The R2 explained by the predictor alone, i.e., in a model containing only that predictor.

##Average.share - The R2 explained by the predictor jointly with other predictors, calculated as the average improvement in R2 across all possible models where that predictor is added. This is the primary metric for individual predictor importance.

##Individual - The total contribution of the predictor to the model Rm. This is the sum of its Unique contribution and its Average.share contribution: Unique+Average.share.

##I.perc(%)	- The percentage contribution of the predictor to the total Rm. Calculated as:
#Individual/Rm2 ×100%.

R2_partition_results_df=R2_partition_results[[2]] %>% 
  as_data_frame() %>% 
  rownames_to_column() %>% 
  rename(predictors=1,
         Unique =2,
         Average.share=3,
         Individual=4,
         `I.perc(%)`=5) %>% 
  mutate(predictors=case_when(predictors%in%"1"~"Year of first record",
                              predictors%in%"2"~"CPI",
                              predictors%in%"3"~"Sources",
                              predictors%in%"4"~"Cumulative intensity",
                              predictors%in%"5"~"Year",
                              predictors%in%"6"~"SST")) %>% 
  # 👇 ADD THIS PART TO SUM DOWN THE COLUMNS AND ADD A NEW ROW
  bind_rows(
    summarise(., 
      predictors = "", # Set the predictor name to an empty string (" ")
      across(
        .cols = c(Unique, Average.share, Individual, `I.perc(%)`), # Specify the columns to sum
        .fns = ~sum(., na.rm = TRUE) # Apply the sum function
      )
    )
  )
  
write.csv(R2_partition_results_df,"R2_partition_results_df_binomial_spatial_sst.csv",row.names = F)
```

Extract predictions
```{r}
test_sp_dt_time_sst_mon_eff_gg_cint=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("annual_c_int [0:183 by =1]"))

test_sp_dt_time_sst_mon_eff_gg_yr=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("year_of_mhws [1987:2020 by =1]"))

test_sp_dt_time_sst_mon_eff_gg_sst=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("median_sst [14:24 by = 0.5]"))

test_sp_dt_time_sst_mon_eff_gg_s_sst=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("max_s_sst [23.7:31.1 by = 0.5]"))

test_sp_dt_time_sst_mon_eff_gg_source=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("sources [0:6 by = 0.1]"))

# test_sp_dt_time_sst_mon_eff_gg_source=ggeffects::ggpredict(
#   test_sp_dt_time_sst_mon_eff,
#   terms=c("sources_pred [0.1:2 by = 0.01]"))

test_sp_dt_time_sst_mon_eff_gg_cpi=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("CPI [-0.09:0.09 by = 0.001]"))

test_sp_dt_time_sst_mon_eff_gg_yfr=ggeffects::ggpredict(
  test_sp_dt_time_sst_mon_eff,
  terms=c("first_rec_year [1924:2020 by = 1]"))

write_rds(test_sp_dt_time_sst_mon_eff_gg_cint,
          "test_sp_dt_time_sst_mon_eff_gg_cint.rds")
write_rds(test_sp_dt_time_sst_mon_eff_gg_yr,
          "test_sp_dt_time_sst_mon_eff_gg_yr.rds")
write_rds(test_sp_dt_time_sst_mon_eff_gg_sst,
          "test_sp_dt_time_sst_mon_eff_gg_sst.rds")
write_rds(test_sp_dt_time_sst_mon_eff_gg_source,
          "test_sp_dt_time_sst_mon_eff_gg_source.rds")
write_rds(test_sp_dt_time_sst_mon_eff_gg_cpi,
          "test_sp_dt_time_sst_mon_eff_gg_cpi.rds")
write_rds(test_sp_dt_time_sst_mon_eff_gg_yfr,
          "test_sp_dt_time_sst_mon_eff_gg_yfr.rds")
write_rds(test_sp_dt_time_sst_mon_eff_gg_s_sst,
          "test_sp_dt_time_sst_mon_eff_gg_s_sst.rds")
```

####Plot model
```{r}
#c_intensity
test_sp_dt_time_sst_mon_eff_cint_p=ggplot(filtered_data_before_first_record_dt_sst_meff_2020)+
  geom_point(aes(x=annual_c_int,y=records),
             alpha=0.3,
             color="grey")+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_cint,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_cint,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="Cumulative intensity (°C-days)",
       y="Record status",
       title="C) MHWs")+
  theme_bw()+
  theme(axis.title.y = element_blank())


#c_intensity zoom
test_sp_dt_time_sst_mon_eff_cint_p_zoom=ggplot(filtered_data_before_first_record_dt_sst_meff_2020)+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_cint,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_cint,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="Cumulative intensity (°C-days)",
       y="Record status")+
  theme_bw()+
  theme(axis.title.y = element_blank())+
coord_cartesian(ylim=c(0,.1))+
  scale_y_continuous(breaks = seq(0,0.1,by=0.05))


#embed cint
embeded_cint <- test_sp_dt_time_sst_mon_eff_cint_p +
  annotation_custom(ggplotGrob(test_sp_dt_time_sst_mon_eff_cint_p_zoom),
                    xmin = 50, xmax = 150,
                    ymin = 0.1, ymax = 0.9)


#years
test_sp_dt_time_sst_mon_eff_yr_p=ggplot(filtered_data_before_first_record_dt_sst_meff_2020)+
  geom_point(aes(x=year_of_mhws,y=records),
             alpha=0.3,
             color="grey")+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_yr,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_yr,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="Year",
       y="Record status",
       title="A) Time")+
  theme_bw()+
  theme(axis.title.y = element_blank())


#SST
test_sp_dt_time_sst_mon_eff_sst_p=ggplot(filtered_data_before_first_record_dt_sst_meff_2020)+
  geom_point(aes(x=median_sst,y=records),
             alpha=0.3,
             color="grey")+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_sst,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_sst,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="SST (°C)",
       y="Record status",
       title="B) SST")+
  theme_bw()+
  theme(axis.title.y = element_blank())

#SST zoom
test_sp_dt_time_sst_mon_eff_sst_p_zoom=ggplot(filtered_data_before_first_record_dt_sst_meff_2020)+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_sst,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_sst,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="SST (°C)",
       y="Record status")+
  theme_bw()+
  coord_cartesian(ylim=c(0,.05))+
  scale_y_continuous(breaks = seq(0,0.05,by=0.025))+
  theme(axis.title = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA))

#embed sst
embeded_sst <- test_sp_dt_time_sst_mon_eff_sst_p +
  annotation_custom(ggplotGrob(test_sp_dt_time_sst_mon_eff_sst_p_zoom),
                    xmin = 15, xmax = 23,
                    ymin = 0.1, ymax = 0.9)+
  theme(axis.title.y = element_blank())

```

```{r}
test_sp_dt_time_sst_mon_eff_grid=gridExtra::grid.arrange(left="Probability of record occurrence",
  test_sp_dt_time_sst_mon_eff_yr_p,
                        embeded_sst,
                        test_sp_dt_time_sst_mon_eff_cint_p,
                        ncol = 2,
                        nrow=2,
  layout_matrix = rbind(c(1, 1,1,1), # Row 1: P1 and P2
                        c(2,2,3,3)))


ggsave(filename ="test_sp_dt_time_sst_mon_eff_grid.png" ,plot =test_sp_dt_time_sst_mon_eff_grid ,device = "jpeg",units = "cm",width = 20,height = 15)
```

Plot non-focal
```{r}
#Source
test_sp_dt_time_sst_mon_eff_source_p=ggplot(filtered_data_before_first_record_dt_sst_meff)+
  geom_point(aes(x=sources,y=records),
             alpha=0.3,
             color="grey")+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_source,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_source,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="Sources",
       y="Record status",
       title="A) Sources")+
  theme_bw()+
  theme(axis.title.y = element_blank())

#CPI
test_sp_dt_time_sst_mon_eff_cpi_p=ggplot(filtered_data_before_first_record_dt_sst_meff)+
  geom_point(aes(x=CPI,y=records),
             alpha=0.3,
             color="grey")+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_cpi,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_cpi,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="CPI (m/s)",
       y="Probability of record occurrence",
       title="B) Current Propagation Index")+
  theme_bw()+
  theme(axis.title.y = element_blank())
test_sp_dt_time_sst_mon_eff_cpi_p

#Year of first record
test_sp_dt_time_sst_mon_eff_yfr_p=ggplot(filtered_data_before_first_record_dt_sst_meff)+
  geom_point(aes(x=first_rec_year,y=records),
             alpha=0.3,
             color="grey")+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_yfr,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_yfr,
              aes(x=x,y=predicted),
            linewidth=1)+
  labs(x="Year of first record",
       y="Probability of record occurrence",
       title="C) Year of first record")+
  theme_bw()+
  theme(axis.title.y = element_blank())+
    scale_x_continuous(breaks = seq(1920,2020,by=20))
test_sp_dt_time_sst_mon_eff_yfr_p

#YFR zoom
test_sp_dt_time_sst_mon_eff_yfr_p_zoom=ggplot(filtered_data_before_first_record_dt_sst_meff)+
  geom_ribbon(data=test_sp_dt_time_sst_mon_eff_gg_yfr,
              aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="#E61316")+
  geom_line(data=test_sp_dt_time_sst_mon_eff_gg_yfr,
              aes(x=x,y=predicted),
            linewidth=1)+
  theme_bw()+
  coord_cartesian(ylim=c(0,.08))+
  scale_y_continuous(breaks = seq(0,0.08,by=0.04))+
  scale_x_continuous(breaks = seq(1920,2020,by=20))+
  theme(axis.title = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA))

#embed yfr
embeded_yfr <- test_sp_dt_time_sst_mon_eff_yfr_p +
  annotation_custom(ggplotGrob(test_sp_dt_time_sst_mon_eff_yfr_p_zoom),
                    xmin = 1940, xmax = 2000,
                    ymin = 0.1, ymax = 0.9)+
  theme(axis.title.y = element_blank())
```

Non climatic panel
```{r}
test_sp_dt_time_sst_mon_eff_grid_nc=gridExtra::grid.arrange(
  left="Probability of record occurrence",
  test_sp_dt_time_sst_mon_eff_source_p,
  test_sp_dt_time_sst_mon_eff_cpi_p,
  embeded_yfr,
  layout_matrix = rbind(c(1), # Row 1: P1 and P2
                        c(2),
                        c(3)))


ggsave(filename ="test_sp_dt_time_sst_mon_eff_grid_nc.png" ,plot =test_sp_dt_time_sst_mon_eff_grid_nc ,device = "jpeg",units = "cm",width = 15,height = 20)
```

###Plot used records
```{r}
# 2. Convert H3 IDs to Spatial Polygons (sf object)
used_records_sf <- filtered_data_before_first_record_dt_sst_meff %>%
  group_by(h3_id) %>%
  summarise(sum_records=sum(records)) %>% 
  # Pass the entire data frame and specify the H3 column name
  h3jsr::cell_to_polygon(
    input = .,               # '.' refers to the incoming data frame
    simple = FALSE           # Recommended for a tidy output
  ) %>%
  # The output is now a proper sf object, so st_set_crs works
  sf::st_set_crs(4326)

# 3. Get Mediterranean Land Boundaries for Context
# Function from rnaturalearth
world <-rnaturalearth:: ne_countries(scale = "large", returnclass = "sf")

# 4. Create the ggplot map
records_map <- ggplot() +
    # Plot the H3 Hexagons (your data)
  geom_sf(
    data = used_records_sf,
    # Fills the hexagons based on the 'sources' count
    aes(fill = sum_records), 
    color = "black",     
    linewidth = 0.1,     
    alpha = 0.8          
  ) +
  # Add land boundaries first 
  geom_sf(data = world, fill = "antiquewhite", color = "black", linewidth = 0.1) +
  # Customize the color scale (from ggplot2 / viridis package, which tidyverse uses)
  scale_fill_viridis_c(
    name = "Records",
    option = "turbo"
  ) +
  # Set the map extent to focus on the Mediterranean Sea
  coord_sf(
    xlim = c(-6, 38),   
    ylim = c(26, 48),  
    expand = FALSE
  ) +
  # Apply a clean theme and labels
  labs(
    title = "A) First records within the temoral range",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw() +
  theme(
# --- CHANGE IS HERE ---
 # Moves the legend inside the plot using (x, y) coordinates
legend.position = c(0.22, 0.17),
# Optional: Makes the legend background transparent
legend.background = element_rect(fill = "transparent", color = NA),
legend.key.size = unit(0.25, "cm"),
panel.grid.major = element_line(color = "skyblue", linewidth = 0.2),
    panel.background = element_rect(fill = "skyblue", color = NA) 
  )
records_map
ggsave(filename ="records_map.png" ,plot =records_map ,device = "jpeg",units = "cm",width = 25,height = 12)
```

sst over time
```{r}
sst_time=filtered_data_before_first_record_dt_sst_meff %>% 
  group_by(h3_id) %>% 
  distinct(h3_id,year_of_mhws,median_sst)
sst_time_p=ggplot(sst_time)+
  geom_point(aes(x=year_of_mhws,y=median_sst,color=median_sst),
             alpha=0.3)+
  geom_boxplot(aes(x=year_of_mhws,y=median_sst,group=year_of_mhws),
               alpha=0.2,notch=T,outliers = F)+
  geom_smooth(aes(x=year_of_mhws,y=median_sst),
              formula = y~x,method = "lm",se = F,color="black")+
  scale_color_viridis_c(
    name = "SST",
    option = "turbo"
  ) +
  labs(x="Years",y="Median SST (°C)",
       title="B) SST trend")+
  guides(color=F)+
  theme_bw()
sst_time_p
ggsave(filename ="sst_time_p.png" ,plot =sst_time_p ,device = "jpeg",units = "cm",width = 20,height = 12)
```

Cumulative intensity over time
```{r}
c_int_time=filtered_data_before_first_record_dt_sst_meff %>% 
  group_by(h3_id) %>% 
  distinct(h3_id,year_of_mhws,annual_c_int)

c_int_time_p=ggplot(c_int_time)+
  geom_point(aes(x=year_of_mhws,y=annual_c_int,color=annual_c_int),
             alpha=0.3)+
  geom_boxplot(aes(x=year_of_mhws,y=annual_c_int,group=year_of_mhws),
               alpha=0.2,notch=F,outliers = F)+
  scale_color_viridis_c(
    name = "Cumulative intensity",
    option = "turbo"
  ) +
  labs(x="Years",y="Cumulative intensity (°C-days)",
       title="C) MHWs detrended")+
  guides(color=F)+
  theme_bw()#+
  # theme(legend.position=c(0.85, 0.8),
  #       legend.background = element_blank(),
  #       legend.key.size = unit(0.4, 'cm'))
c_int_time_p
ggsave(filename ="c_int_time_p.png" ,plot =c_int_time_p ,device = "jpeg",units = "cm",width = 20,height = 12)
```

Into grid
```{r}
intro_grid=gridExtra::grid.arrange(
  records_map,
  sst_time_p,
  c_int_time_p,
  ncol = 2,
  nrow=2,
  layout_matrix = rbind(c(1,1,1,1,1,1), # Row 1: P1 and P2
                        c(1,1,1,1,1,1),
                        c(1,1,1,1,1,1),
                        c(2,2,2,3,3,3),
                        c(2,2,2,3,3,3)))

intro_grid
ggsave(filename ="intro_grid.png" ,plot =intro_grid ,device = "jpeg",units = "cm",width = 20,height = 20)
```


##dredge
```{r}
# Dredge searches all possible combinations of fixed effects.
# It automatically keeps the random effects ((1|h3_id) + (1|species)) in ALL models.

# 1.1 Set MuMIn's options to prevent NA's from failing the model
# This is crucial for dredge()
options(na.action = "na.fail")

# It also generates the null model (intercept only) and the global model (all fixed effects).
cat("\nPerforming Model Dredging (comparing all 64 models)...\n")
model_selection_table <-MuMIn::dredge(
  test_sp_dt_time_sst_mon_eff,
  rank = "AICc" # Use AIC corrected for small sample sizes
)

write_rds(model_selection_table,"model_selection_table_cut_2020.rds")
write.csv(model_selection_table %>% as_tibble(),"model_selection_table_cut_2020.csv",row.names = F)

# ==============================================================================
# STEP 3: MANUAL FILTERING AND AIC WEIGHT CALCULATION
# ==============================================================================
cat("\n-- Filtering and Recalculating Weights for Converged Models --\n")

# 3.1 Filter out all models that failed to converge (where AICc is NA)
valid_models <- subset(model_selection_table, !is.na(AICc))

# 3.2 Recalculate Delta AICc based ONLY on the converged models' best score
# This ensures delta=0 corresponds to the best CONVERGED model
min_AICc <- min(valid_models$AICc)
valid_models$delta <- valid_models$AICc - min_AICc

valid_models=valid_models %>% 
  as.data.frame()
valid_models$weight=as.numeric(valid_models$weight)
# 3.3 Manually Calculate Akaike Weights (Formula: exp(-0.5 * delta) / sum(exp(-0.5 * delta)))
# This is equivalent to MuMIn's weight() function, applied to the subset
valid_models=valid_models %>% 
  mutate(weight=exp(-0.5 * valid_models$delta) / sum(exp(-0.5 * valid_models$delta))) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))

# Sort the valid models by the newly calculated AICc rank
valid_models <- valid_models[order(valid_models$AICc), ]
print(failed_models)

#write.csv(valid_models,"valid_selected_models.csv",row.names = F)

# ==============================================================================
# STEP 4: OUTPUT RESULTS
# ==============================================================================

cat("\n=======================================================\n")
cat(paste0(" MODEL SELECTION TABLE (N=", nrow(valid_models), " Converged Models) "))
cat("\n=======================================================\n")

# Print the top 10 *converged* models
print(head(valid_models, n = 10))

cat("\nInterpretation:\n")
cat(" - The AICc, delta, and weight columns are now only calculated for the 46 converged models.\n")
cat(" - delta: The difference in AICc between a model and the best CONVERGED model (delta=0).\n")
cat(" - weight: The Akaike weight, showing the probability of being the best model in this set.\n")

# Calculate variable importance (Sum of weights for each predictor)
# We must use the original sw() function on the VALID models
importance_ranking <- MuMIn::sw(valid_models)
cat("\n=======================================================\n")
cat("             RELATIVE VARIABLE IMPORTANCE              \n")
cat("=======================================================\n")
# Importance close to 1 indicates the variable is in all or most top-ranking models.
print(importance_ranking)

# Reset global options
options(na.action = "na.omit")
```

###dredge binomial <2020
```{r}
options(na.action = "na.fail")

# It also generates the null model (intercept only) and the global model (all fixed effects).
cat("\nPerforming Model Dredging (comparing all 64 models)...\n")
model_selection_table_binomial <-MuMIn::dredge(
  test_sp_dt_time_sst_mon_eff,
  rank = "AICc" # Use AIC corrected for small sample sizes
)

write_rds(model_selection_table_binomial,"model_selection_table_binomial_s_sst.rds")
write.csv(model_selection_table_binomial %>% as_tibble(),"model_selection_table_cut_2020_binomial_s_sst.csv",row.names = F)

options(na.action = "na.omit")

# Calculate variable importance (Sum of weights for each predictor)
# We must use the original sw() function on the VALID models
importance_ranking <- MuMIn::sw(model_selection_table_binomial)
importance_ranking %>% 
  as_tibble()
```


###test with DCIp
```{r}
test_sp_dt_time_sst_mon_eff_dcip=glmmTMB(
  data = filtered_data_before_first_record_dt_sst_meff,
  formula = records~
    scale(first_rec_year)+
    scale(DCI_P)+
    scale(sources)+
    scale(annual_c_int)+
    scale(year_of_mhws)+
    scale(median_sst)+
    (1|h3_id)+
    (1|species),
  family = "betabinomial")
test_sp_dt_time_sst_mon_eff_dcip %>% summary
```

####dredge - dcip
```{r}

# Dredge searches all possible combinations of fixed effects.
# It automatically keeps the random effects ((1|h3_id) + (1|species)) in ALL models.

# 1.1 Set MuMIn's options to prevent NA's from failing the model
# This is crucial for dredge()
options(na.action = "na.fail")

# It also generates the null model (intercept only) and the global model (all fixed effects).
cat("\nPerforming Model Dredging (comparing all 64 models)...\n")
model_selection_table_dcip <-MuMIn::dredge(
  test_sp_dt_time_sst_mon_eff_dcip,
  rank = "AICc" # Use AIC corrected for small sample sizes
)
model_selection_table_dcip
write_rds(model_selection_table_dcip,"model_selection_table_dcip.rds")
write.csv(model_selection_table_dcip %>% as_tibble(),"model_selection_table_dcip.csv",row.names = F)

# ==============================================================================
# STEP 3: MANUAL FILTERING AND AIC WEIGHT CALCULATION
# ==============================================================================
cat("\n-- Filtering and Recalculating Weights for Converged Models --\n")

# 3.1 Filter out all models that failed to converge (where AICc is NA)
valid_models_dcip <- subset(model_selection_table_dcip, !is.na(AICc))

# 3.2 Recalculate Delta AICc based ONLY on the converged models' best score
# This ensures delta=0 corresponds to the best CONVERGED model
min_AICc <- min(valid_models_dcip$AICc)
valid_models_dcip$delta <- valid_models_dcip$AICc - min_AICc

valid_models_dcip=valid_models_dcip %>% 
  as.data.frame()
valid_models_dcip$weight=as.numeric(valid_models_dcip$weight)
# 3.3 Manually Calculate Akaike Weights (Formula: exp(-0.5 * delta) / sum(exp(-0.5 * delta)))
# This is equivalent to MuMIn's weight() function, applied to the subset
valid_models_dcip=valid_models_dcip %>% 
  mutate(weight=exp(-0.5 * valid_models_dcip$delta) / sum(exp(-0.5 * valid_models_dcip$delta))) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))

# Sort the valid models by the newly calculated AICc rank
valid_models_dcip <- valid_models_dcip[order(valid_models_dcip$AICc), ]
print(failed_models)

#write.csv(valid_models_dcip,"valid_selected_models_dcip.csv",row.names = F)

# ==============================================================================
# STEP 4: OUTPUT RESULTS
# ==============================================================================

cat("\n=======================================================\n")
cat(paste0(" MODEL SELECTION TABLE (N=", nrow(valid_models), " Converged Models) "))
cat("\n=======================================================\n")

# Print the top 10 *converged* models
print(head(valid_models_dcip, n = 10))

cat("\nInterpretation:\n")
cat(" - The AICc, delta, and weight columns are now only calculated for the 46 converged models.\n")
cat(" - delta: The difference in AICc between a model and the best CONVERGED model (delta=0).\n")
cat(" - weight: The Akaike weight, showing the probability of being the best model in this set.\n")

# Calculate variable importance (Sum of weights for each predictor)
# We must use the original sw() function on the VALID models
importance_ranking <- MuMIn::sw(valid_models_dcip)
cat("\n=======================================================\n")
cat("             RELATIVE VARIABLE IMPORTANCE              \n")
cat("=======================================================\n")
# Importance close to 1 indicates the variable is in all or most top-ranking models.
print(importance_ranking)

# Reset global options
options(na.action = "na.omit")
```

